

Upon installation the user is able to access models using the "standard" 
table of models stored in `allometric_models`. This is a reduced set of 
searchable information that is suitable for most use-cases. More advanced users
may prefer to search a more complete set of information to find their models. 
This vignette covers what types of extra information can be searched and some 
techniques for searching these extra data. Interested users would do well to 
read [examples...]

The object `allometric_models_ext` is also available to the user for searching,
but represents a more complex data structure. This complexity arises from 
`column-lists` which are columns of a data frame that contain lists (or vectors) as their 
elements. A motivating example is the `family_name` column, which contains 
character vectors for each model that contain the family names (i.e., last names)
of all contributing authors.

Let's say we want to search for any model with `'Hann'` as a contributing author.
This can be done with some elegant use of `dplyr::filter` and `purrr::map_lgl`.
Before beginning, it is helpful to know that `purrr::map_lgl` returns a list
of TRUE/FALSE values as it "maps" over a list of input.

Consider the vector `auth <- c('Hann', 'Weiskettel')`, I could check if this vector 
contains `'Hann'` using the expression `'Hann' %in% auth`. Knowing this, let's
take a look at filtering out all models containing `Hann` as an author:

```{r eval=F}
dplyr::filter(
    allometric_models,
    purrr::map_lgl(family_names, ~ 'Hann' %in% .)
)
```

We can even check for models that contain certain types of data requirements. 
For example, the following block finds diameter-height models, specifically 
models that use diameter outside bark at breast height as the *only* covariate.
The utility here is obvious, since many inventories are vastly limited by their
available measurements.

```{r eval=F}
dplyr::filter(
    allometric_models,
    measure == 'height',
    purrr::map_lgl(covt_names, ~ length(.)==1 & .[[1]] == 'dsob'),
)
```

Breaking this down, we have the first condition `measure=='height'` selecting
only models concerned with heights as a response variable. The second line 
maps over each element of the `covt_names` column, which is a character vector.
The `.` represents a given character vector for that row. First, we ensure that
the vector is only one element in size using `length(.)==1`, then we ensure that
the first (and only) element of this vector is equal to `'dsob'`, (diameter
outside bark at breast height).
