---
title: "Describing a Model with Descriptors"
vignette: >
  %\VignetteIndexEntry{Describing a Model with Descriptors}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include = F}
library(allometric)
```

Descriptors are sets of information that describe a model. The user can think 
of descriptors as a nested list of helpful information for finding and
specifying models. For example, Brackett (1977) is a technical report that 
fits 24 cubic volume models for a wide variety of tree species, age classes,
and size classes. Descriptors answer the question: how would a user navigate
models stored in `allometric` in a way that is consistent with the publication?
We will take a look at Brackett (1977) as a motivating example.

In Brackett (1977) some species, like Douglas-fir, have multiple models that
are fit to specific geographic regions and age classes. Let's take a look at
the descriptors of one of those models

```{r}
brackett_acer
```

The last section of the output specifies all descriptors for this model, 
including `country`, `region`, `family`, `genus`, `species`, `geographic_region`
and `age_class`. These descriptors are enough for a user to identify the
model within the publication.

In this case, the first five descriptors (`country` through `species`) are 
referred to as "search descriptors", which are descriptors propagated to the
`allometric_models` dataframe. Contributors are encouraged to defined these 
descriptors whenever possible. The remaining two descriptors are non-standard
descriptors specific to this publication, and use verbage from the publication
to assist users in finding the correct model. Contributors are free to name
these descriptors in a way that makes sense for the publication.

## Specifying Descriptors at Different Levels

`allometric` operates at three levels arranged in a hierarchy. There are
publications, model sets within publications and models within model sets.
For convenience, contributors can define descriptors at any level, with the
understanding that all descriptors defined at that level apply uniformly to 
all models lower in the hierarchy.

For example, the source code for the Brackett model installation specifies 
`country` and `region` at the publication-level.

```{r eval=F}
brackett_1977 <- Publication(
  citation = bracket_1977_citation,
  descriptors = list(
    country = "US",
    region = "US-WA"
  )
)
```

This means that **all** models in `brackett_1977` are constructed with data only
from the state of Washington in the United States. The `FixedEffectsSet`,
`MixedEffectsSet` and their respective `FixedEffectsModel` and `MixedEffectsModel`
classes all have a `descriptors` argument that can be used for specifying
descriptors at these levels. Refer to the [Installing a Model](https://allometric.github.io/allometric/articles/installing_a_model.html)
vignette for more examples at these levels.
