---
title: "Describing a Model with Descriptors"
vignette: >
  %\VignetteIndexEntry{Describing a Model with Descriptors}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include = F}
library(allometric)
```

Descriptors are sets of information that describe a model. The user can think 
of descriptors as a nested list of helpful information for finding and
specifying models. For example, Brackett (1977) is a technical report that 
fits 24 cubic volume models for a wide variety of tree species, age classes,
and size classes. Descriptors answer the question: how would a user navigate
models stored in `allometric` in a way that is consistent with the publication?
We will take a look at Brackett (1977) as a motivating example.

In Brackett (1977) some species, like Douglas-fir, have multiple models that
are fit to specific geographic regions and age classes. Let's take a look at
the descriptors of one of those models

```{r}
brackett_acer
```

The last section of the output specifies all descriptors for this model, 
including `country`, `region`, `family`, `genus`, `species`, `geographic_region`
and `age_class`. These descriptors are enough for a user to identify the
model within the publication.

In this case, the first five descriptors (`country` through `species`) are 
referred to as "search descriptors", which are descriptors propagated to the
`allometric_models` dataframe. Contributors are encouraged to defined these 
descriptors whenever possible. The remaining two descriptors are non-standard
descriptors specific to this publication, and use verbage from the publication
to assist users in finding the correct model. Contributors are free to name
these descriptors in a way that makes sense for the publication.
