---
title: "Common Inventory Use Cases"
vignette: >
  %\VignetteIndexEntry{Common Inventory Use Cases}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include=F}
knitr::opts_chunk$set(eval = FALSE)
```

This vignette covers several common forest inventory use cases that can be 
covered by `allometric`

## Making Predictions of Volume for Multiple Species with Models of the Same Form

This vignette demonstrates a simple forest inventory example using `allometric`
using plot data obtained from the Forest Inventory and Analysis (FIA) program.
An example dataset is included in the package and can be loaded using 
`data(fia_trees)`. Our objective is to calculate a plot-level volumes for 
every plot in this dataset.

Allometric predictions can be made in any manner of ways according to the 
user's preference. We will use the suite of `tidyverse` packages to manipulate
our data, but this is merely the author's preference. `allometric` is capable
of producing predictions using base R if preferred. For this example, we will
use a set of models that all have the same functional form. That is, the
required covariates are identitcal across models, and are in the same order.
For a more advanced case, refer to the following section.

### Preparing the Data

In reading `?fia_trees` the user will notice that all tree species are encoded
using the FIA `SPCD` numbering system. The first task is to prepare a data frame
that will translate these codes into taxonomic groupings.

```{r}
library(tibble)

target_species <- tibble(
  SPCD = c(202, 263, 15, 122),
  genus = c('Pseudotsuga', 'Tsuga', 'Abies', 'Pinus'),
  species = c('menziesii', 'heterophylla', 'concolor', 'ponderosa')
)
```

The `target_species` dataframe serves two purposes. First, it allows us to link
`genus` and `species` information to `fia_trees`. Second, it allows us to 
easily filter out allometric models that are not related to these types of
trees.

### Finding Appropriate Models

The user will note that `fia_trees` is composed entirely of trees from the state
of Oregon in the United States. Using `dplyr` and `purrr` we can efficiently 
filter out unnecessary models from the `allometric_models` table. Refer to the
`?allometric_models` help page for more information.

For this example, we need stem volume models for our species of interest that,
preferably, were fit using data in the state of Oregon.

```{r message = FALSE}
library(allometric)
library(dplyr)
library(purrr)

install_models()
allometric_models <- load_models()

stemvol_mods <- allometric_models %>%
  filter(
    model_type == "stem volume"
  )

nrow(stemvol_mods)
```

We can see that `stemvol_mods` contains a huge amount of stem volume models.
That is because `allometric` is global in scope. Most of these models are not
relavent for our needs. Next, we will drill down and select models that were 
developed with data from `US-OR` (the state of Oregon, United States).

```{r}
stemvol_or_mods <- stemvol_mods %>%
  filter(map_lgl(region, ~ 'US-OR' %in% .))

nrow(stemvol_or_mods)
```

Here we see all stem volume models available fit using data from the state of
Oregon. Still, the amount of models is very high. Next, we will select only our
target species. First, we use the `unnest_taxa()` function, that will append
`family`, `genus`, and `species` columns, which will facilitate merging the
target species later.

```{r}
stemvol_or_unnested <- stemvol_or_mods %>%
  unnest_taxa()
```

```{r}
stemvol_or_spc_mods <- stemvol_or_unnested %>%
  inner_join(target_species, by=c('genus', 'species'))

nrow(stemvol_or_spc_mods)
```

Finally, something a bit more digestible. Even still, we have multiple models
per species. The last step is to ensure that each of these remaining models are
compatible with our available data. `fia_trees` only contains `DIA`, the 
diameter at breast height, and `HT` the total height of the tree. In `allometric`
these covariates are standardized as `dsob` and `hst` respectively. See the
[Variable Naming System](https://allometric.github.io/allometric/articles/variable_naming_system.html)
for more information. Using some clever filtering, we can find which models are
compatible with this set of covariates.

```{r}
stemvol_or_spc_dia_ht_mods <- stemvol_or_spc_mods %>%
  filter(map_lgl(covt_name, ~ length(.) == 2 & all(c('dsob', 'hst') %in% .)))
```

The call to `map_lgl` answers the question, "which models contain only two 
covariates that are named `dsob` and `hst`"? Only a handful of models remain.

```{r}
stemvol_or_spc_dia_ht_mods
```

At this stage, we only need to determine a single model for Pseudotsuga
menziesii. We will decide to select only those models from `poudel_2019`, a
recent study that contains the volume models for the species we need.

```{r}
final_set <- stemvol_or_spc_dia_ht_mods %>% filter(pub_id == "poudel_2019") %>%
  select(SPCD, model)
final_set
```

At last we have all four species attached to one allometric model, and we can
proceed with prediction. If the above steps seem cumbersome, fear not, the
"Storing an Internal Set of Equations for Routine Use" section resolves these
issues.

### Making Volume Predictions

To make a volume prediction correctly, we must know the order of the arguments
to the allometric function. Inspecting the first model in our set reveals this
order.

```{r}
final_set %>% mutate(call = map_chr(model, ~model_call(.)))
```

The `model_call` function shows that `vsa = f(dsob, hst)` is the allometric 
function call for each model we have specified, which means that the
diameter of the stem outside bark at breast height (`dsob`) is the
first argument and the total height of the stem (`hst`) is the second argument.

Using `final_set`, we can easily merge the models to our tree-list and make
predictions. Predictions made in this manner are done using the `predict_allo`
function.

```{r}
tree_vols <- final_set %>%
  left_join(fia_trees, by = "SPCD") %>%
  mutate(vol_m3 = predict_allo(model, DIA * 2.54, HT * 0.3048)) %>%
  mutate(vol_cuft = units::set_units(vol_m3, "ft^3"))

head(tree_vols)
```

`predict_allo` takes each model in the `model` column and applies it to the
covariates, `DIA` and `HT` assigned in the function call. Aggregating to the
plot-level we obtain

```{r}
tree_vols %>%
  group_by(PLOT) %>%
  summarise(vol_ac = sum(vol_cuft * TPA_UNADJ))
```

Note that all models in `final_set` take centimeters for diameter and meters for
height. This can be checked by printing the model summary

```{r}
final_set$model[[1]]
```

It is always important to check the model summary before implementing a model
in an analysis. Indeed, users should carefully read the publication source as
well to be absolutely sure the model is implemented appropriately.

## Making Predictions of Volume for Multiple Species with Models of Different Forms

In many cases, allometric model inputs will differ in the required covariates
and, potentially, their order. This complicates the use of `predict_allo`, which
assumes that the covariates are identical and of the same order for each model
used in the analysis.

For this example we will use a different set of models from the same
publication, Poudel et al. (2019) to calculate tree biomass (`bt`) for the same
set of trees.

### Finding Appropriate Models

In this case we have a publication in mind, and can filter `allometric_models`
directly to obtain the models we want. We want a set of tree-level biomass
models from `poudel_2019` that use `dsob` or `hst` as covariates. We obtain

```{r}
biomass_models <- allometric_models %>%
  filter(
    pub_id == "poudel_2019",
    model_type == "tree biomass",
    map_lgl(covt_name, ~(any(c("dsob", "hst") %in% .) & length(.) <= 2))
  ) %>%
  mutate(call = map_chr(model, ~model_call(.))) %>%
  unnest_taxa() %>%
  right_join(target_species, by = c("genus", "species"))
```

We can see that the model for *Abies concolor* takes only a `dsob` argument but
the remaining models take `dsob` and `hst` arguments. This renders the direct
use of `predict_allo` on `biomass_models` infeasible.

Instead we must engage in some clever programming and break our dataframe of
models into call groups, use `predict_allo` for each group, and stitch the
prediction frame back together for our final result.

### Making Biomass Predictions with Different Model Forms

First, we need to make a function that will take each call group and its
covariate key. This allows us to implement different prediction logic for each
call group using a tree of `if` statements. We obtain:

```{r}
call_group_predict <- function(data, covt_key) {
  if(identical(covt_key[[1]][[1]], c("dsob"))) {
    data$agb_kg <- predict_allo(data$model, data$DIA * 2.54)
  } else if (identical(covt_key[[1]][[1]], c("dsob", "hst"))) {
    data$agb_kg <- predict_allo(data$model, data$DIA * 2.54, data$HT * 0.3048)
  }

  data
}
```

Then, we simply merge the models to the tree list, group by the `covt_name`
column, and utilize `dplyr::group_map()` to apply our `call_group_predict`
function to each group of trees.

```{r}
tree_agbs <- biomass_models %>%
  left_join(fia_trees, by = "SPCD") %>%
  group_by(covt_name) %>%
  group_map(call_group_predict) %>%
  bind_rows()

head(tree_agbs)
```

Finally, we can aggregate the `tree_agbs` to the plot-level as before.

```{r}
agb_plot <- tree_agbs %>%
  group_by(PLOT) %>%
  summarise(agb_ac = sum(agb_kg * TPA_UNADJ))

head(agb_plot)
```

## Storing an Internal Set of Equations for Routine Use

It is unlikely that most organizations need all models stored in
`allometric_models` for routine use. For example, the Forest Inventory and 
Analysis (FIA) program has specified sets of allometric models over the years
that are used to produce cubic volumes and biomasses for the national FIA
database. Subsets of the `allometric_models` table can be prepared by
analysts and stored locally as `RDS` files, which can then be used in routine
analysis or even as components of other software packages.

For example, we can save the model table we created in the last section

```{r eval=F}
saveRDS(final_set, './my_directory/final_set.RDS')
```

It can be reloaded and used readily in other software

```{r eval=F}
readRDS('./my_directory/final_set.RDS')
```

If you will be using other `allometric` functionality in your software, (e.g.,
`predict_allo`) it will be necessary to run `library(allometric)` to gain access
to those functions.