---
title: "Mixed-Effects Models"
---

Mixed-effects models are common in the allometric modeling literature, and
allow "calibration" of models at different levels of aggregation, such as 
field plots and stands. `allometric` supports mixed effects modeling with the 
`MixedEffectsModel` class. However, due to the highly technical nature of this topic,
this vignette highlights a few important concerns that authors should be aware
of  before proceeding with a mixed-effects model implementation.

## Navigating Published Mixed Effects Models

Implementing published mixed-effects models is, strictly speaking, 
possible in `allometric`. However, predictions for groups with
random parameters (e.g., field plots or stands) require knowledge of the 
estimated variance-covariance matrix of the random parameters themselves. As
well, partial derivatives of the model with respect to the
fixed-effects parameters are needed. These two components are not always
explicitly provided in publications. 

Because `allometric` directly cites publications and implements models from 
them, it is best practice to only implement what is explicitly published, 
rather than derive un-published portions of the models.
[Temesgen et al. (2008)](https://cdnsciencepub.com/doi/full/10.1139/X07-104?casa_token=cS6uvsVOt9wAAAAA%3AQDIa9ji2K78DzvfT87xh1qJy0IaqFam2kh4xiaRN13AU0rXEHwQW243Gkfbek1SQ8RFmGqkrvklw)
is one such example where the derivatives are evalutated (p. 557), while
[Sharma and Parton (2007)][https://www.sciencedirect.com/science/article/pii/S0378112707004069?casa_token=kzkeRCmkwKIAAAAA:UgOjEJCXiwN7IDkyLdxzOUXheeaKEQk01z2YaQABnG15nEC4Hya_hQ22c0XMyUPzHiCEBXHV5g]
is an example where the derivatives are not evaluated (Section 2.4). Using this
approach, the mixed-effects models in Temesgen et al. (2008) are appropriate for
`allometric` but  the mixed-effects models in Sharma and Parton (2007) are not,
since the  final model form is not strictly provided in the publication.

Alternatively, predictions from mixed-effects models can be generated using the
fixed-effects estimates only, which are often reported. In fact, many studies
will assess the performance of a mixed-effects model in this way, and the
utility of using the fixed-effects estimates is obvious when observations from a
new "target group" are not available. However, this activity does deviate from
the purest intent of the model, and should be explicitly flagged in `allometric`
by installing a `MixedEffectsModel` class with the argument `fixed_only = TRUE`
to denote that only the fixed-effects are generating the predictions. This
argument is propagated to further documentation and model summaries to alert
users to this special case.

Using this approach, we can still include the mixed-effects models in
papers like Sharma and Parton (2007) using only the fixed-effects estimates,
but we flag the model with the `fixed_only` argument to maintain clarity. Users
with the technical skill needed to program mixed-effects models are free to 
implement predicted random effects themselves.

## Implementing a Mixed Effects Model

```{r include=FALSE}
library(allometric)
```

Mixed-effects models require an additional function to generate predictions for
the random effects called `predict_ranef`. Similar to `predict_fn`, this 
function takes covariates as arguments and generates predictions. However, the
user must also provide a `newdata`, which represents data from a new group of
observations, such as a field plot or stand. This dataframe must contain columns
with the same covariates used in `predict_ranef` as well as new observed values
of the response variables.

`predict_ranef` must return a list with the names of the predicted random
effects, which are then propagated to `predict_fn`. Let us examine on example 
from Temesgen et al. (2008). We begin with the `predict_fn`:

```{r}
predict_fn <- function(dsob) {
  1.37 + (beta_0 + b_i) * (1 - exp(beta_1 * dsob)^beta_2)
}
```

Here we see a typical `predict_fn`. In this case, we know ahead of time that the
authors are using `b_i` as the only random effect (by reading the paper,
of course). `beta_0`, `beta_1` and `beta_2` are the fixed-effects. So, we need
a function that will produce predictions of `b_i` given a set of new
observations. We define `predict_ranef`

```{r}
predict_ranef <- function(hst, dsob) {
  z <- hst - beta_0 * (1 - exp(beta_1 * dsob)^beta_2)
  Id <- diag(length(hst))
  b_i <- sigma_sq_b * t(z) %*% solve(sigma_sq_epsilon * Id * z %*% t(z)) %*%
    (hst - beta_0 * (1 - exp(beta_1 * dsob))^beta_2)
  list(b_i = as.numeric(b_i))
}
```

While somewhat complicated, this function is merely producing predictions of 
`b_i` (third line) after calculating some pre-requisite values. Then, the 
random effect is stored in a named list and returned (last line).

Finally, we prepare the entire model:

```{r}
mem <- MixedEffectsModel(
  response_unit = list(
    hst = units::as_units("m")
  ),
  covariate_units = list(
    dsob = units::as_units("cm")
  ),
  parameters = list(
    beta_0 = 40.4218,
    beta_1 = -0.0276,
    beta_2 = 0.936,
    sigma_sq_b = 6.544,
    sigma_sq_epsilon = 2.693
  ),
  predict_ranef = predict_ranef,
  predict_fn = predict_fn
)
```

Note that all parameters we need for `predict_ranef` and `predict_fn` are 
defined in the `parameters` argument. Typically, these require estimates of 
the variance components in addition to the betas.