---
title: "Variable Naming System"
---

```{r include=F}
library(allometric)
```

## Variable Naming Structure

The variable naming system is a standard approach for classifying allometric
response variables and covariates. In the allometric modeling literature, 
a diverse set of names are available for many similar types of variables.
For example, one may wish to refer to the volume of the main stem of a tree, 
the volume of the stump, the diameter of a stump, the diameter at breast height,
etc. Analysts may also wish to refer to different spatial scales, such as plot-
or stand-level variables.

The basic structure of a variable name is:

```
[increment_flag_][measure][component][...modifiers][_scale_flag]
```

The `[increment_flag_]` and `[_scale_flag]` portions are optional flags that
allow a user to specify if the variable is an increment (e.g., diameter increment),
and the scale flag allows a user to specify plot- or stand-level variables.
If these flags are not specified, then the variable is assumed to be a
non-increment tree-level variable.

`[measure]` and `[component]` are both single characters that encode the type
of measurement (e.g., diameter, height, etc.) and the portion of the tree 
the variable is referred to (e.g., stem, foliage, etc.), respectively. These
are followed by optional modifiers that allow further specification of the
variable.

The `get_variable_def` function allows the user to search for existing variable
definitions. For example, we can examine a common variable name, `"dsob"`

```{r}
get_variable_def("dsob")
```

and find that it is the diameter of the stem, outside bark at breast height. We
see that the measure is `"d"` for diameter, the component is `"s"` for stem
and two modifiers are used (bark and height modifiers) with values of `"o"` and
`"b"` respectively.

If we want to refer to a diameter outside bark increment, simply add the 
increment flag `"i_"`:

```{r}
get_variable_def("i_dsob")
```

If we want to refer to a plot-level diameter outside bark (e.g., the mean
diameter of trees on a plot), add a scale flag `"_p"` as a suffix.

```{r}
get_variable_def("dsob_p")
```

there is also a scale flag for stand-level variables, `"_s"`.

## Measures

All variable names begin with a "measure". In the example above, we searched 
the naming system for `d` measures, but others are available. There are
`r nrow(measure_defs)` established measures:

```{r}
measure_defs
```

We recognize that the use of "measure" here is somewhat loose (e.g., diameter
and height are both _measures_ of length). However, the variable naming system
is meant to preserve at least some connection to commonly used terms for 
variables in allometric modeling.

## Components

All variable names then have a second character that specifies the tree 
component. There are `r nrow(component_defs)` established components:

```{r}
component_defs
```

Mixing these first two character codes provides an expressive system to refer to 
measures and components. For example, the biomass of foliage, if one were 
interested in it, would begin `bf`, the volume of the stem would be `vs`,
and the diameter of the stump would be `du`.

## Measure-Component Modifiers

Within a given measure-component pair, a number of useful modifiers are 
available. These are specific to the measure-component pair. For example, an
astute modeler may be wondering, "how do I refer to diameter at breast height
outside bark"? These modifiers enable this level of specificity. 

### Diameter of Stem (`ds`) Modifiers

Some of the most common allometric variables are that of stem diameter, which
arises in taper modeling, merchantable heights and, of course, diameter at 
breast height. `ds` modifiers enable a user to specify these broad range of 
variables. The available `ds` variables are:

```{r}
get_variable_def("ds")
```

### Height of Stem (`hs`) Modifiers

```{r}
get_variable_def("hs")
```

### Volume of Stem (`vs`) Modifiers

```{r}
get_variable_def("vs")
```

### Biomass (`b`) Components and Modifiers

```{r}
get_variable_def("b")
```