---
title: "Variable Naming System"
---

```{r include=F}
library(allometric)
options(width=1000)
```

The variable naming system is a standard approach for classifying allometric
response variables and covariates. In the allometric modeling literature, 
a diverse set of names are available for many similar types of variables.
For example, one may wish to refer to the volume of the main stem of a tree, 
the volume of the stump, the diameter of a stump, the diameter at breast height,
etc.

The variable naming system, while somewhat verbose and strict, is worth the 
effort. It allows `allometric` to efficiently categorize many allometric models
in a standardized structure, which facilitates searching and documentation.
Users can utilize the `get_variable_def()` function to search for partial 
matches of variables in the naming system. For example, to find all definitions
for diameter variables use:

```{r}
get_variable_def("d")
```

This lists all variables in the naming system that have `"d"` as their first
`search_str` character. We can inspect, e.g., `dsob`

```{r}
get_variable_def("dsob")
```

and find that it is the diameter of the stem, outside bark at breast height.

This vignette is not comprehensive for all variables stored in the package, 
but should cover the most common cases. `get_variable_def()` will contain all
possible variables.

## Measures

All variable names begin with a "measure". In the example above, we searched 
the naming system for `d` measures, but others are available. There are
`r nrow(measure_defs)` established measures:

```{r}
measure_defs
```

We recognize that the use of "measure" here is somewhat loose (e.g., diameter
and height are both _measures_ of length). However, the variable naming system
is meant to preserve at least some connection to commonly used terms for 
variables in allometric modeling.

## Components

All variable names then have a second character that specifies the tree 
component. There are `r nrow(component_defs)` established components:

```{r}
component_defs
```

Mixing these first two character codes provides an expressive system to refer to 
measures and components. For example, the biomass of foliage, if one were 
interested in it, would begin `bf`, the volume of the stem would be `vs`,
and the diameter of the stump would be `du`.

## Measure-Component Modifiers

Within a given measure-component pair, a number of useful modifiers are 
available. These are specific to the measure-component pair. For example, an
astute modeler may be wondering, "how do I refer to diameter at breast height
outside bark"? These modifiers enable this level of specificity. 

### Diameter of Stem (`ds`) Modifiers

Some of the most common allometric variables are that of stem diameter, which
arises in taper modeling, merchantable heights and, of course, diameter at 
breast height. `ds` modifiers enable a user to specify these broad range of 
variables. The available `ds` variables are:

```{r}
get_variable_def("ds")
```

### Height of Stem (`hs`) Modifiers

```{r}
get_variable_def("hs")
```

### Volume of Stem (`vs`) Modifiers

```{r}
get_variable_def("vs")
```

### Biomass (`b`) Components and Modifiers

```{r}
get_variable_def("b")
```