---
title: "Variable Naming System"
vignette: >
  %\VignetteIndexEntry{Variable Naming System}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc} 
---

```{r include=F}
library(allometric)
options(width = 10000)
```

## Variable Naming Structure

The variable naming system is a standard approach for naming allometric
response variables and covariates. In the allometric modeling literature, 
a diverse set of names are available for many similar types of variables.
For example, the frequently used acronym "DBH" is used to refer to the diameter
at breast height, but DBH can refer to different measures of the stem, such as
the diameter inside bark at breast height, or the diameter outside bark at
breast height.

The variable naming system provides a standard set of names that clarify the use
of the variable while remaining succinct and easy to write. In the variable
naming system, the diameter outside bark at breast height is referred to as
**dsob**. Let us inspect what this name means further as a motivating example.

The first two characters in any variable name are referred to as the "measure"
and "component", respectively. In the case of **dsob** the **d** refers to the
diameter and the **s** refers to the stem. The remaining characters are 
modifiers that clarify the context of the measure-component pair. In this case
**o** refers to outside bark and **b** refers to breast height.

The core structure of a variable name is:

```
[measure][component][...modifiers]
```

The `get_variable_def` function allows the user to search for existing variable
definitions. For example, we can examine our example variable `"dsob"`

```{r}
get_variable_def("dsob")$description
```

and find that it is the diameter of the stem, outside bark at breast height. We
can view the full definition below, which includes a complete breakdown of the
measure, component and modifiers of the variable name:

```{r}
get_variable_def("dsob")
```

In the following sections we will discuss how variable names are structured.

## Measures

All variable names begin with a "measure". Measures describe the dimensional
aspects of what is being measured for the variable. In the example above, we
searched the naming system for `d` (diameter) measures, but others are
available. There are `r nrow(get_measure_defs())` established measures:

```{r}
get_measure_defs()
```

We recognize that the use of "measure" here is somewhat loose (e.g., diameter
and height are both _measures_ of length). However, the variable naming system
is meant to preserve at least some connection to commonly used terms for 
variables in allometric modeling.

## Components

All variable names then have a second character that specifies the tree 
component. Components describe the part of the tree that is being measured.
There are `r nrow(get_component_defs())` established components:

```{r}
get_component_defs()
```

Mixing these first two character codes provides an expressive system to refer to 
measures and components. For example, the biomass of foliage, if one were 
interested in it, would begin `bf`, the volume of the stem would be `vs`,
and the diameter of the stump would be `du`.

## Measure-Component Modifiers

Within a given measure-component pair, a number of useful modifiers are 
available. These are specific to the measure-component pair. We highlight a few
useful groups of modifiers below. For further information users can call
`get_variable_def` with the measure-component pair of their choice.

### Diameter of Stem (`ds`) Modifiers

Some of the most common allometric variables are that of stem diameter, which
arise in taper modeling, merchantable heights and, of course, diameter at 
breast height. `ds` modifiers enable a user to specify these broad range of 
variables. The available `ds` variables are:

```{r}
get_variable_def("ds")
```

### Height of Stem (`hs`) Modifiers

Another common group of variables are those that describe stem height, such as
the total height of the stem, site index, height of the stem to crown base,
and many others.

```{r}
get_variable_def("hs")
```

### Volume of Stem (`vs`) Modifiers

Stem volume models are among the most common models available in `allometric`,
including volume of the stem inside or outside bark, volume of the stem above
breast height, and many others.

```{r}
get_variable_def("vs")
```

### Biomass (`b`) Components and Modifiers

Biomass models compose a vast portion of the allometric modeling literature.
All biomass models are stored under the `b` measure, and include component
biomass models (e.g., branch, foliage, etc.) as well as whole-tree biomass
models.

```{r}
get_variable_def("b")
```

## Increment and Scale Flags

Two additional flags can be added to any variable name to indicate increment
models (e.g., height increment) or change the spatial scale of the variable
(e.g., plot-level biomass).

Increment flags exist as the prefix `i_`. For example, we can modify the `dsob`
variable to describe diameter outside bark increment via `i_dsob`:

```{r}
get_variable_def("i_dsob")
```

Scale flags exist as a suffix, with two suffixes available, one for the plot
scale, `_p`, and one for the stand scale, `_s`. For example, we can refer to
the stand-level basal area as

```{r}
get_variable_def("gs_s")
```

We can even combine increment and scale flags to describe, for example,
stand-level growth variables, e.g., a stand-level basal area increment would
be expressed as `i_gs_s`

```{r}
get_variable_def("i_gs_s")
```