---
title: "`allometric` v2.0.0 Developer Notes"
vignette: >
  %\VignetteIndexEntry{Frequently Asked Questions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r include=FALSE}
devtools::load_all()
models <- load_models()
```

Recently, `allometric` was updated to v2.0.0. The move from v1 to v2 does not
necessarily indicate the delivery of new features. Rather, it indicates that
substantial changes were made to the package that prevent back-compatibility
with v1. Hence, we encourage all readers to update their `allometric`
installations and models by running:

```{r eval=FALSE}
devtools::install_github("allometric/allometric")

library(allometric)
install_models(redownload = TRUE)
```

Largely, these updates only effect the back-end of `allometric`, e.g., those
functions that allow users to contribute models to our [repository](https://github.com/allometric/models).
The one exception are the `Taxa` and `Taxon` classes, described below.

## Introducing the `Taxon` and `Taxa` Classes

Prior versions of `allometric` tracked species information for a model using the
`family`, `genus`, and `species` descriptor fields. These were insufficient for
several reasons. First, a model can belong to multiple taxonomic specifications
(e.g., Douglas-fir and Ponderosa pine together). In `allometric` v1, this was
achieved by duplicating the model and assigning the one species to each copy,
which is ineffecient for storage purposes. Second, there were no validity checks
used to ensure that the taxonomic specification was valid. This means a model
could just have a `species` field defined, but no `genus` or `family`, which
causes data quality issues.

In `allometric` v2, taxonomic specifications must use the `Taxa` and `Taxon`
classes, which enforce stricter checks for taxonomic validity. `Taxa` is
essentially a special type of list that only contains `Taxon` objects. `Taxon`
is an object that contains one taxonomic hierarchy.

For example:

```{r}
pp <- Taxon(family = "Pinaceae", genus = "Pinus", species = "ponderosa")
wwp <- Taxon(family = "Pinaceae", genus = "Pinus", species = "monticola")

my_taxa <- Taxa(pp, wwp)
```

specifies a `Taxa` containing two types of trees, Ponderosa pine and Western
white pine. Some useful methods have been implemented for `Taxa` and `Taxon`.
For example, we can do common set operations on `Taxa`:

```{r}
"Pinus" %in% my_taxa

pp %in% my_taxa
```

We can do the same for `Taxon`. A few examples:

```{r}
"Pinus" %in% pp
"ponderosa" %in% wwp

pp == pp
pp == wwp
```

These methods facilitate searching, described in the next section.

### Searching Taxa in a Model Table

One of the most common operations users will perform is searching through a
model table to find models of a given family, genus, or species. Using the above
operations, filtering on the `taxa` column is similar to working with any other
hierarchical column in a model table. Readers should already be familiar with
the `purrr` examples on [this page](https://allometric.org/reference/load_models.html).

```{r eval=FALSE}
ponderosa_models <- models %>%
  dplyr::filter(
    purrr::map_lgl( # A map over each model's taxa
      taxa,
      ~ pp %in% .
    )
  )
```

The structure of this example is exactly the same as any other hierarchical
search done in `allometric`. We filter the set of models using
`dplyr::filter()` on the `taxa` column. For each `Taxa` we encounter,
represented by `.`, we check if our Ponderosa pine `Taxon` is inside. If yes,
the function `~ pp %in% .` returns true, and the row is retained. Please
refer to [this page](https://allometric.org/reference/load_models.html) for
further examples.

### Usage when Writing Models

The `taxa` descriptor is a list of `Taxon` elements. Here is an example for a
height-diameter model that applies to both Douglas-fir and Western hemlock.

```{r}
my_model <- FixedEffectsModel(
  response = list(hst = units::as_units("ft")),
  covariates = list(dsob = units::as_units("in")),
  parameters = list(a = 1),
  predict_fn = function(dsob) {a * dsob},
  descriptors = list(
    taxa = list(
      Taxon(
        family = "Pinaceae",
        genus = "Pseudotsuga",
        species = "menziesii"
      ),
      Taxon(
        family = "Pinaceae",
        genus = "Tsuga",
        species = "heterophylla"
      )
    ),
    country = "US",
    region = "US-OR"
  )
)
```

We can see that the value of the `taxa` descriptor is a list of an arbitrary
number of `Taxon` calls. Calls to `Taxon` need not specify all three values of
the hierarchy. Instead, we can specify shallower layers only:

```{r}
Taxon(family = "Pinaceae")
```

but we cannot specify deeper layers without specifying the shallower layers
first:

```{r error = TRUE}
Taxon(genus = "Pinus")
```

### Usage when Writing Model Sets

For model sets we have added a helper function `aggregate_taxa()`, which will
aggregate a row of a parameter CSV file across the `family`, `genus` and
`species` columns into a new column called `taxa`. For example:

```{r}
my_parameter_frame <- tibble::tibble(
  family = c("Pinaceae", "Betulacae"),
  parameter_1 = c(1, 2)
)

my_parameter_frame %>% aggregate_taxa()
```

This allows back-compatibility with previously written parameter files from
v1, while enforcing the taxa structure.

## JSON Representations of Models

Models can now be converted to and from JSON. Of all of the new updates, this
is the one I am most excited about, because it enables models to be stored in
a remotely hosted database. The to/from JSON conversions are a necessary
pre-requisite for this database, which will yield many forthcoming features
including:

1. A publicly available API for searching/requesting model information.
2. An interactive web viewer for the models
3. Entrypoints for other languages, such as Python, to access models and use
them in analysis.

In short, the remotely hosted database will become the centralized location of
the models, and users will interact with this database exclusively in the
future. For now, users must still install models locally. We expect to deliver
the database in 2024.

Point 3 is especially exciting, and savvy Python package developers are
encouraged to email me at bfrank70@gmail.com for some colloboration ideas.

## `covariate_units` and `response_unit` Refactoring

The `covariate_units` and `response_unit` arguments were refactored to
`covariates` and `response`, respectively. Only users that contribute models
to `allometric` will be impacted by this change. For example, a
`FixedEffectsModel` is now made by the following:

```{r eval=F}
FixedEffectsModel(
  response = list(
    vsia = units::as_units("ft^3")
  ),
  covariates = list(
    dsob = units::as_units("in")
  ),
  ...
)
```

This is easier to write and remember.
